---
title: "Midterm Review"
---

```{r, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, comment=NA, fig.show="hide")
library(tidyverse)
```

## Chapter 3: Probability
Normal distribution 

$X \sim \text{N}(8,3)$

$\text{P}(X < 10)$ 


```{r}
dfs <- tibble(
  x=seq(-5, 25, length=500), 
  `f(x)` = dnorm(x,  mean=8, sd=3)
  ) 

gp <- ggplot(dfs) + 
  aes(x = x, y = `f(x)`) + 
  geom_area(alpha = 0.4) 

gp + 
  geom_area(
    data = dfs |> 
      filter(x < 10),
    fill="coral1"
    ) 
```

```{r}
pnorm(q=10, mean=8, sd=3)
```

Poisson Distribution
The number of parasites in a host is a Poisson random variable with mean 3.5. What is the probability that there will be at least one parasite in the host? 

$Z \sim \text{P}(\lambda = 3.5)$

```{r}
1 - dpois(x=0, lambda=3.5)
```
Binomial Distribution

A microbiologist conducts an experiment to create a recombinant strain of bacteria that is resistant to penicillin. She plates out the bacteria on a plate, and picks out 10 colonies. She knows that the probability of successfully creating a recombinant is 0.15.
Given $X \sim \text{Bin}(n = 10, p = 0.15)$, what is $P(x>0)$

```{r}
1 - dbinom(x=0, size=10, prob=0.15)
```

## Chapter 4: Statistical Inference

```{r}
url1 <- "http://www.massey.ac.nz/~anhsmith/data/rangitikei.RData"
download.file(url = url1, destfile = "rangitikei.RData")
load("rangitikei.RData")
```

```{r}
# ggplot style
library(ggplot2)
p1 <- ggplot(rangitikei, mapping = aes(people))+
  geom_histogram(aes(y=..density..), bins=10)

p1 + stat_function(fun = dnorm, args=list(mean=mean(rangitikei$people), sd=sd(rangitikei$people)), geom = "line")
```

```{r}
# Old style
qqnorm(rangitikei$people)
qqline(rangitikei$people)

# ggplot style
ggplot(rangitikei, aes(sample=people))+stat_qq()+stat_qq_line()

```

```{r}
# test for normality
shapiro.test(rangitikei$people)

ks.test(rangitikei$people, "pnorm")  # ties

```
What type of test does each of the following code chucks specify? State a null and alternative hypothesis for each. Interpret results.

```{r}
t.test(rangitikei$people, mu=100)
# Null true mean of people is = 100
# alternative is the true mean of people is not = 100
```

```{r}
t.test(rangitikei$people, mu=100, alternative="greater")
# Null true mean of people is less than or equal to 100
# alternative is the true mean of people is greater than 100
```

```{r}
t.test(rangitikei$people~factor(rangitikei$time))
t.test(people~factor(time), data = rangitikei, var.equal = TRUE)
```

Transformations

According to this boxcox figure what is the most appropriate transformation to try first.
```{r}
library(lindia)
gg_boxcox(lm(rangitikei$people~1))
```
How would you perform this transformation for a ttest?

```{r}
rep_sq_people <- (-1-sqrt(rangitikei$people))
t.test(rep_sq_people, mu=100, alternative="greater")
# Null true mean of people is less than or equal to 100
# alternative is the true mean of people is greater than 100
```
## Chapter 5: Tabulated Counts

# Exercise 5.2 {-}

The genetic information of an organism is stored in its Deoxyribonucleic acid (DNA). DNA is a double stranded helix made up of four different nucleotides. These nucleotides differ in which of the four bases Adenine (A), Guanine (G), Cytosine (C), or Thymine (T) they contain. A simple pattern that we may want to detect in a DNA sequence is that of the nucleotide at position `i+1` based on the nucleotide at position `i`. The nucleotide positional data collected by a researcher in a particular case is given in the following table:

| `i\(i+1)` | A   | C   | G   | T   |
|-----------|-----|-----|-----|-----|
| A         | 622 | 316 | 328 | 536 |
| C         | 428 | 262 | 204 | 306 |
| G         | 354 | 294 | 174 | 266 |
| T         | 396 | 330 | 382 | 648 |

Perform a test of association and then obtain the symmetric plot.

```{r}
tabledata <- data.frame(
  A = c(622, 428, 354, 396),
  C = c(316, 262, 294, 330),
  G = c(328, 204, 174, 382),
  T = c(536, 306, 266, 648), 
  row.names = c("A", "C", "G", "T")
  )
```

```{r, fig.show='hide', results='hide'}
chisq.test(tabledata)$exp
chisq.test(tabledata)
chisq.test(tabledata, simulate.p.value = T)

library(MASS)
corresp(tabledata)
plot(corresp(tabledata, nf=2))
abline(v=0)
abline(h=0)
#or
library(FactoMineR)
CA(tabledata)

```

## Test questions

## 21 
n=5

mean = 158

var = 20

```{r}
#95% ci for the mean
n <- 5
mu <- 158
sd <- sqrt(20)

SE <- sd / sqrt(n)
SE

```
$$ 
\bar x \pm t \times \text{se}
$$

The $t$ value is the 0.975 quantile of the $t$ distribution with the degrees of freedom given by $n-1$. 

```{r}
qt(p = 0.975, df = n - 1)
```


```{r}
mu + (2.77*SE)
mu - (2.77*SE)
```

```{r}
mu + (qt(p = 0.975, df = n - 1)*SE)
```






```{r}
n = 5
s = sqrt(12)

se = s / sqrt(n) 
se
```


$$ 
\bar x \pm t \times \text{se}
$$
The $t$ value is the 0.975 quantile of the $t$ distribution with the degrees of freedom given by $n-1$. 

```{r}
qt(p = 0.975, df = n - 1)
```
So, the sample mean and confidence interval is:

```{r}
( xbar <- 161 )
( xbar - qt(p = 0.975, df = n - 1) * se )
( xbar + qt(p = 0.975, df = n - 1) * se )
```

## 22 Two fertilizers
n=23
stand n=12, mu=101, var=18
new n=11, mu=124, var=14

pooled variance
```{r}
# w1= (n1-1)/
```


## 23
```{r}
new_fert <- c(124.8, 118.5, 128.8, 117.8, 124.2, 122.3, 114.5, 120.7, 123.9, 119.1, 121.5)
stand_fert <- c(106.4, 105.5, 103.8, 97.7, 96.5, 91.4, 97.7, 99.6, 97, 92.3, 103.9, 102)

mean(new_fert - stand_fert)

mean(new_fert) - mean(stand_fert)

# y+t x S/sqrt(n)
# 
# n=11
```

